Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-04-14 manuscript/manuscript_file.Rmd
added discussion of NO3 and temperature partial dependence plots

### N~2~O undersaturation extent  
In this study we present the first national-scale survey of N~2~O concentrations and emissions from lakes, reservoirs, and ponds, hereafter referred to as waterbodies, in the conterminous U.S. (CONUS). A generalized random-tessellation survey design (Stevens and Olsen, 2004) was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[weird, I have 1090, but NLA17 report says 1,005?]** sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week (Figure 1). Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate (See NLA website for all data?). `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks (Figure 1). We modeled the data using a bayesion hierchical approach and predicted the dissolved N~2~O concentration for all 224,916 CONUS waterbodies in the target population. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`), the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), and >65% of the waterbodies in every ecoregion were undersaturated (Table 1).

@@

#### Controls on N~2~O saturation
N~2~O saturation was related to waterbody size in several ecoregions, but the direction of the relationship varied.  For example, N~2~O saturation increased with waterbody size in the Northern Plains, Upper Midwest, Northern Appalachians, and Western Mountain ecoregions, but decreased with increasing size in the Coastal and Southern Plains (Figure 3A).  The inconsistent relationship between N~2~O saturation and waterbody size may be attributable differences in nitrate concentrations across ecoregions.  The model identified an interaction between waterbody surface area and nitrate concentration (expressed categorically) where N~2~O concentration decreases with increasing lake size at high NO3 concentrations, but increases, or remains constant at low nitrate concentrations. THIS IS BECAUSE THERE CAN'T BE MUCH N2O BIOGEOCHEMISTRY WITHOUT INORGANIC N LAYING AROUND.

@@

Regardless of the direction of the relationship between N~2~O saturation and waterbody size, N~2~O saturation deviates from equilibrium most strongly in the smallest waterbody size class.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases linearly with increasing waterbody size class (Figure 3B).  Waterbody size can affect dissolved gas concentration through several mechanisms.  In lakes with low volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column.  Since water volume to benthic surface area ratios tend to scale positively with lake surface area, benthic N~2~O biogeochemistry will have a greater affect on dissolved N~2~O concentration in small lakes, regardless of whether the sediments function as a net sink or source of N~2~O **[see Wetzel for references]**.  Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes **[need a citation for N2O ~ temperature relationship]**.  Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (Beaulieu et al N2O Harsha Lake and USACE N2O papers).

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequlibrim in smaller waterbodies.  Air-water gas exchange will vent excess N~2~O from surface waters or replenish depleted N~2~O in undersaturated waterbodies by facilitating the invasion of atmospheric N~2~O.  Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium.  One of the primary drivers of air-water gas exchange in lentic waters is wind speed (Vachon and Prairie reference).  Wind speeds to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation.  Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.

@@

>>>>

### N~2~O undersaturation extent  
In this study we present the first national-scale survey of N~2~O concentrations and emissions from lakes, reservoirs, and ponds, hereafter referred to as waterbodies, in the conterminous U.S. (CONUS). A generalized random-tessellation survey design (Stevens and Olsen, 2004) was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[weird, I have 1090, but NLA17 report says 1,005?]** sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week (Figure 1). Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate (See NLA website for all data?). `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks (Figure 1). We modeled the data using a bayesion hierchical approach and predicted the dissolved N~2~O concentration for all 224,916 CONUS waterbodies in the target population **[where did I get 224,916, Roy has 465,897]**. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`), the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), and >65% of the waterbodies in every ecoregion were undersaturated (Table 1).

@@

#### Controls on N~2~O
Dissolved N~2~O concentration increased with NO~3~ concentration **[need to pick a figure]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the  sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ (Knowles 1982; Richardson et al. 2009).  The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column.  Incomplete denitrification often occurs when NO~3~ is relatively abundant, likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification (Koike and Hattori, 1975;Richardson et al., 2009). When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~ thereby functioning as an N~2~O sink.

It is also possible that the correlation between N2O and NO3 reflects the simultaneous production of these solutes via nitrification.  Nitrification, an aerobic process which can produce N2O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ **[citation]**.  Nitrification might be expected to be more prevalent in the predominantly well oxygenated surface waters where the N~2~O samples were collected than denitrification, which is an anaeorbic process.  However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column **[citation]** and can be transported from aoxic sediments to the water surface via advection and diffusion (Delsontro no longer a paradox).  Thus the N~2~O - NO3 relationship may be reflective of some combination of N2O production/consumption via denitrification and N2O production via nitrification.

The sensitivity of N2O to NO3 increased with water temperature **[need to pick a figure]**.  This is consistent with N2O production and consumption occurring through a temperature sensitive biological process.  This relationship is apparent at high NO3 concentrations, where nitrogen biogeochemistry is a net source of N2O, and at low NO3 concentrations where nitrogen biogeochemistry is a net sink for N2O.  Thus NO3 concentration dictates whether a water body is a source or sink of N2O and elevated water temperature magnifies the effect.

**[N2O vs NO3 with different lines representing quantiles of the temperature distribution]**

#### Controls on N~2~O saturation **[this heading is to organize the .Rmd, not intended to be retained in final MS]**

While NO3 availability determined whether N biogeochemistry was a source or sink of N2O and water temperature affected the rate of N2O production or consumption, waterbody morphology determined the extent to which N2O production or consumption affected dissolved N~2~O concentration at the air-water interface.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases linearly with increasing waterbody size class (Figure 3A). Waterbody size can affect dissolved gas concentration through several mechanisms. In lakes with low volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column.  Since water volume to benthic surface area ratios tend to scale positively with lake surface area, benthic N~2~O biogeochemistry will have a greater affect on dissolved N~2~O concentration in small lakes, regardless of whether the sediments function as a net sink or source of N~2~O **[see Wetzel for references]**.  Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes **[need a citation for N2O ~ temperature relationship]**.  Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (Beaulieu et al N2O Harsha Lake and USACE N2O papers).

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequlibrim in smaller waterbodies.  Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies.  Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium.  One of the primary drivers of air-water gas exchange in lentic waters is wind speed (Vachon and Prairie reference).  Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation.  Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.

The relationship between N~2~O and waterbody size is also evident in a few ecoregions (Figure 3B).  For example, mean N2O saturation ratio is greater than 1 for all size classed in the Coastal Plains, consistent with the high NO3 concentrations in this ecoregion **[is this true?]**, but decreases with increasing lake size, consistent the effect of waterbody morphology on N2O production and gas exchange.  By contrast, mean N2O saturation is less than one across all size classes in the low NO3 northern plains, but again, N2O saturation moves toward equilibrium with increasing lake size.  These relationships are evident to lesser degrees in other ecoregions, likely dependent on the distribution on NO3 and water temperature across lake classes.  **[I'm beginning to think that Fig. 3B unnecessarily complicates an otherwise pretty clean story.  Maybe we just show Fig. 3B?]**

@@

@@

# EMISSION RATES
